\bf Estimating the number of clusters using cross-validation

### 2019-06-29 JCGS-R3/cvclust.tex
fix self-consistency, wold refs

With many clustering methods, including the popular $k$-means clustering
procedure, the user must specify $k$, the number of clusters
\citep{jain2010data}.  One popular ad-hoc device for selecting the number of
clusters is to use an analogue of the principal components scree plot: plot
the within-cluster dispersion $W_k$, as a function of the number of clusters
$k$, looking for an ``elbow'' in the plot.  This approach is simple and often
performs well, but it requires subjective judgment as to where the elbow is
located, and as we demonstrate in Appendix~\ref{sec:elbow-fail}, the approach
can easily fail.  In this report, we propose a new method to choose the number
of clusters automatically.

@@

The procedure we propose in this report is based on a form of
cross-validation, and it is adaptive to the characteristics of the data
distribution. The essential idea is to devise a way to measure a form of
internal prediction error associated each choose of $k$, and then choose the
$k$ with the smallest associated error. We describe this method in detail in
Section~\ref{sec:meth}.  In Section~\ref{sec:self-consistent}, we prove that
our method is self-consistent in certain situations, meaning that in the
absence of noise, it recovers the correct $k$. Then, in
Section~\ref{sec:theory-gaussian}, we
analyze the performance of our method in the presence of Gaussian noise. The
theoretical analysis shows that the performance of our method degrades in the
presence of correlated noise; to fix this, we propose a correction for
correlated noise in Section~\ref{sec:corr-correct}. In
Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation}, we
demonstrate that our method is competitive with other state-of-the-art
procedures in both simulated and real data sets. Then, in
Section~\ref{sec:application-yeast}, we apply our method to a Yeast cell cycle
dataset. We conclude with a short discussion in Section~\ref{sec:discussion}.

@@

%% TODO: appendix

In Section~\ref{sec:self-consistent}, we prove that this procedure is
self-consistent, in the sense that it recovers the correct value of $k$ in the
absence of noise. Then, in Section~\ref{sec:theory-gaussian}, we analyze some
of the properties of Gabriel cross-validation in the presence of Gaussian
noise.

@@

@@

Proposition~\ref{prop:self-consistency} tells us that the Gabriel cross-validation
method recovers the true number of clusters when the noise is negligible.
While this result gives us some assurance that the procedure is well-behaved,
we can bolster our confidence and gain insight into its workings by analyzing
its behavior in the presence of noise.  We first study the case of a single
cluster in two dimensions with correlated Gaussian noise.

@@

In Appendix~\ref{sec:single-general-dim}, using similar techniques to those used
to prove Proposition~\ref{prop:single-2d}, we derive an analogous result for correlated
Gaussian noise in more than two dimensions. A similar phenomenon holds:
the Gabriel cross-validation method fails when the first principal component
of the $Y$ variables is strongly correlated with a linear combination of the
$X$ variables.

@@

\begin{proof}
There are two clusters: observations from cluster~$1$ are distributed as
\(
  \mathcal{N}(\mu, I)
\)
and observations from cluster~$2$ are distributed as
\(
\mathcal{N}(-\mu, I)
\)
where $\mu = (\muX, \muY)$.  Without loss of generality, $\muX \geq 0$ and
$\muY \geq 0$.  Let $G_i$ be the true cluster of observation~$i$
where, by assumption,
\[
  \Pr(G_i=1) = \Pr(G_i=2) = 1/2.
\]
After applying $k$-means to $\{ Y_i\}_{i=1}^{n}$ with $k=2$, if $n$ is large
enough, then the estimated cluster means $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$
will be close to $\E(Y \mid Y>0)$ and $\E(Y \mid Y < 0)$, with errors of
size $\OhP(n^{-1/2})$. To compute these quantities, let
$(X_1, Y_1) \sim \mathcal{N}(\mu, I)$ and $(X_2, Y_2) \sim \mathcal{N}(-\mu, I)$
be draws from the mixture components, and let $(X,Y)$ be defined such that
$\Pr(X= X_1, Y = Y_1) = \Pr(X= X_2, Y = Y_2) = 1/2$. Then,
\begin{align*}
  \E(Y \mid Y>0)
  &= \E(Y_1 \mid Y_1 > 0) \cdot \Pr(Y = Y_1 \mid Y >0)
   + \E(Y_2 \mid Y_2 > 0) \cdot \Pr(Y = Y_2 \mid Y > 0)
\\
&=  \{ \muY + \varphi(\muY) / \Phi(\muY) \} \cdot \Phi(\muY)
+ [-\muY + \varphi(\muY) / \{ 1 - \Phi(\muY) \}] \cdot \{ 1 - \Phi(\muY) \}
\\
  &= 2\varphi(\muY)+ 2\muY\Phi(\muY)- \muY. 
\end{align*}
In the second line, we have used
Lemma~\ref{lem:truncated-normal-moments} from
Appendix~\ref{app:technical-lemmas} to compute the conditional expectations;
$\varphi()$ and $\Phi()$ are the standard normal density and
cumulative distribution function, respectively. By symmetry,
\begin{align*}
  \E(Y \mid Y<0) &= -\E(Y \mid Y > 0).
\end{align*}

@@

We have stated and proved Proposition~\ref{prop:twoclust} for data in $d = 2$
dimensions, but there is nothing essential about the particulars of the
dimension; we generalize the result to arbitrary dimension in
Appendex~\ref{app:twoclust-anyd}.

@@

The transformation used in step~\ref{step:transform} uses a random orthogonal
matrix~$Q$, which gets applied to the rows of $\dataX$ after 
multiplying by the estimate of $\Sigma^{-1/2}$. We use this random orthogonal
matrix for two reasons. First, it ensures that in expectation, each transformed cluster mean
$Q \Lambda^{-1/2} \Gamma^\T \mu(g)$ for $g = 1, \dotsc, k$ is uniformly
spread across all $P$ features. This ensures that the self-consistency
conditions on the cluster centers enumerated in Proposition~\ref{prop:self-consistency}
are likely to hold. The second reason for multiplying by $Q$ is to spread any
remaining correlation in the noise evenly (in expectation) across all
dimensions. The latter effect follows since if $Z$ is a random vector with
covariance matrix $\Theta$, then $Q Z$ has covariance matrix $Q \Theta Q^\T$,
which has expectation $E(Q \Theta Q^\T) = \tr(\Theta) I$.

@@

In this section, we perform a set of simulations to evaluate the performance
of our proposed method and the associated correlation correction described in
Section~\ref{sec:corr-correct}.  We compare our method with a basket of
competing methods including the Gap statistic
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, the CH-index \citep{calinski1974dendrite}, Hartigan's
statistic \citep{hartigan1975clustering}, the Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
Slope method \citep{fujita2014non} and Bootstrap stability \citep{fang2012selection}. 
We use the default parameter settings for all competing methods, implented
in the following R~packages:
\texttt{cluster} \citep{cluster};
\texttt{cstab} \citep{cstab};
\texttt{e1071} \citep{e1071};
\texttt{fpc} \citep{fpc};
\texttt{MASS} \citep{MASS};
\texttt{mclust} \citep{mclust};
and
\texttt{NbClust} \citep{NbClust}.
For Gabriel cross-validation,
we perform $2$-fold cross-validation on the columns~($p=q$) and $5$-fold
cross-validation on the rows~($m=n/4$).  We also compare with Wold
cross-validation, which we describe in Appendix~\ref{sec:wold-cv}.

>>>>

With many clustering methods, including the popular $k$-means clustering
procedure, the user must specify $k$, the number of clusters
\citep{jain2010data}.  One popular ad-hoc device for selecting the number of
clusters is to use an analogue of the principal components scree plot: plot
the within-cluster dispersion $W_k$, as a function of the number of clusters
$k$, looking for an ``elbow'' in the plot.  This approach is simple and often
performs well, but it requires subjective judgment as to where the elbow is
located, and can easily fail (we demonstrate such a failure in Section~1 of
the supplementary appendix).  In this report, we propose a new method to
choose the number of clusters automatically.

@@

The procedure we propose in this report is based on a form of
cross-validation, and it is adaptive to the characteristics of the data
distribution. The essential idea is to devise a way to measure a form of
internal prediction error associated each choose of $k$, and then choose the
$k$ with the smallest associated error. We describe this method in detail in
Section~\ref{sec:meth}.  Then, in
Section~\ref{sec:theory-gaussian}, we
analyze the performance of our method in the presence of Gaussian noise. The
theoretical analysis shows that the performance of our method degrades in the
presence of correlated noise; to fix this, we propose a correction for
correlated noise in Section~\ref{sec:corr-correct}. In
Sections~\ref{sec:simulations} and~\ref{sec:empirical-validation}, we
demonstrate that our method is competitive with other state-of-the-art
procedures in both simulated and real data sets. Then, in
Section~\ref{sec:application-yeast}, we apply our method to a Yeast cell cycle
dataset. We conclude with a short discussion in Section~\ref{sec:discussion}.

@@

@@

We turn now to theoretical properties of Gabriel cross-validation. In
Section~2 of the supplementary appendix, show that the procedure is
self-consistent, in the sense that it recovers the correct value of $k$ in the
absence of noise.  This gives us some assurance that the procedure is
well-behaved, but we can bolster our confidence and gain insight into its
workings by analyzing its behavior in the presence of noise. To that end, we
devote the present section to analysis of the procedure in the presence of
Gaussian noise.

@@

We first study the case of a single
cluster in two dimensions with correlated Gaussian noise.

@@

In Section~3 of the supplementary appendix,
%\ref{sec:single-general-dim}
using similar techniques to those used
to prove Proposition~\ref{prop:single-2d}, we derive an analogous result for correlated
Gaussian noise in more than two dimensions. A similar phenomenon holds:
the Gabriel cross-validation method fails when the first principal component
of the $Y$ variables is strongly correlated with a linear combination of the
$X$ variables.

@@

\begin{proof}
There are two clusters: observations from cluster~$1$ are distributed as
\(
  \mathcal{N}(\mu, I)
\)
and observations from cluster~$2$ are distributed as
\(
\mathcal{N}(-\mu, I)
\)
where $\mu = (\muX, \muY)$.  Without loss of generality, $\muX \geq 0$ and
$\muY \geq 0$.  Let $G_i$ be the true cluster of observation~$i$
where, by assumption,
\[
  \Pr(G_i=1) = \Pr(G_i=2) = 1/2.
\]
After applying $k$-means to $\{ Y_i\}_{i=1}^{n}$ with $k=2$, if $n$ is large
enough, then the estimated cluster means $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$
will be close to $\E(Y \mid Y>0)$ and $\E(Y \mid Y < 0)$, with errors of
size $\OhP(n^{-1/2})$. To compute these quantities, let
$(X_1, Y_1) \sim \mathcal{N}(\mu, I)$ and $(X_2, Y_2) \sim \mathcal{N}(-\mu, I)$
be draws from the mixture components, and let $(X,Y)$ be defined such that
$\Pr(X= X_1, Y = Y_1) = \Pr(X= X_2, Y = Y_2) = 1/2$. Then,
\begin{align*}
  \E(Y \mid Y>0)
  &= \E(Y_1 \mid Y_1 > 0) \cdot \Pr(Y = Y_1 \mid Y >0)
   + \E(Y_2 \mid Y_2 > 0) \cdot \Pr(Y = Y_2 \mid Y > 0)
\\
&=  \{ \muY + \varphi(\muY) / \Phi(\muY) \} \cdot \Phi(\muY)
+ [-\muY + \varphi(\muY) / \{ 1 - \Phi(\muY) \}] \cdot \{ 1 - \Phi(\muY) \}
\\
  &= 2\varphi(\muY)+ 2\muY\Phi(\muY)- \muY. 
\end{align*}
In the second line, we have used
Lemma~3 from Section~5 of the supplementary appendix
    to compute the conditional expectations;
$\varphi()$ and $\Phi()$ are the standard normal density and
cumulative distribution function, respectively. By symmetry,
\begin{align*}
  \E(Y \mid Y<0) &= -\E(Y \mid Y > 0).
\end{align*}

@@

We have stated and proved Proposition~\ref{prop:twoclust} for data in $d = 2$
dimensions, but there is nothing essential about the particulars of the
dimension; we generalize the result to arbitrary dimension in
Section~4 of the supplementary appendix.
%Appendex~\ref{app:twoclust-anyd}.

@@

The transformation used in step~\ref{step:transform} uses a random orthogonal
matrix~$Q$, which gets applied to the rows of $\dataX$ after 
multiplying by the estimate of $\Sigma^{-1/2}$. We use this random orthogonal
matrix for two reasons. First, it ensures that in expectation, each transformed cluster mean
$Q \Lambda^{-1/2} \Gamma^\T \mu(g)$ for $g = 1, \dotsc, k$ is uniformly
spread across all $P$ features. This ensures that the cluster centers
are well identified.
The second reason for multiplying by $Q$ is to spread any
remaining correlation in the noise evenly (in expectation) across all
dimensions. The latter effect follows since if $Z$ is a random vector with
covariance matrix $\Theta$, then $Q Z$ has covariance matrix $Q \Theta Q^\T$,
which has expectation $E(Q \Theta Q^\T) = \tr(\Theta) I$.

@@

In this section, we perform a set of simulations to evaluate the performance
of our proposed method and the associated correlation correction described in
Section~\ref{sec:corr-correct}.  We compare our method with a basket of
competing methods including the Gap statistic
\citep{tibshirani2001estimating}, Gaussian mixture model-based clustering
\citep{fraley2002model}, the CH-index \citep{calinski1974dendrite}, Hartigan's
statistic \citep{hartigan1975clustering}, the Jump method
\citep{sugar2003finding}, Prediction strength \citep{tibshirani2005cluster},
Slope method \citep{fujita2014non} and Bootstrap stability \citep{fang2012selection}. 
We use the default parameter settings for all competing methods, implented
in the following R~packages:
\texttt{cluster} \citep{cluster};
\texttt{cstab} \citep{cstab};
\texttt{e1071} \citep{e1071};
\texttt{fpc} \citep{fpc};
\texttt{MASS} \citep{MASS};
\texttt{mclust} \citep{mclust};
and
\texttt{NbClust} \citep{NbClust}.
For Gabriel cross-validation,
we perform $2$-fold cross-validation on the columns~($p=q$) and $5$-fold
cross-validation on the rows~($m=n/4$).  We also compare with Wold
cross-validation, which we describe in section~6 of the supplementary
appendix. %Appendix~\ref{sec:wold-cv}.
